---
title: "species-comparison"
author: "Cenaida Largo"
format: html
editor: visual
---

# Overview

This script explores the alignment of the iNaturalist earthworm (Crassiclitellata) taxonomy with:

- **WoRMS** (World Register of Marine Species), the current taxonomy source used by iNaturalist.

- **Misırlıoğlu et al.(2023)**, an updated global checklist of valid earthworm species.

Subspecies are excluded from this analysis.

## Setup

Package versions are tracked with \`renv\`.

```{r}
#| label: setup-packages
#| message: false
#| warning: false

library(here)     #Project relative file paths
library(readr)    #Import data files into R
library(readxl)   #Import excel files into R
library(dplyr)    #Manipulating data
library(stringr)  #Manipulating text
```

## Data sources

### iNaturalist

Download all “active” names from the **iNaturalist Taxonomy DarwinCore Archive** at

<https://www.inaturalist.org/pages/developers> retrieved 05-2026.

Updated monthly.

Notes

Taxon author name and year aren’t stored in iNaturalist.

I submitted a feature request to include the taxa authority in the DWCA taxonomy export but it was denied by the forum moderator.

See previous discussion here: https://forum.inaturalist.org/t/include-author-citations-for-taxa/671.

### WoRMS

**WoRMS taxon list** downloads are available at

<https://www.marinespecies.org/aphia.php?p=download&what=taxlist>

Because the web interface limits exports to 1000 rows, a custom export was requested at

<https://www.marinespecies.org/usersrequest.php>

Requested fields:

- ScientificName

- Authority

- taxonRank

- taxonomicStatus

- Classification

- Accepted taxon information

Make sure to change the environment field from “marine = yes” to “marine = any”.

Contact: info(at)marinespecies.org

### **Misirlioğlu et al 2023**

**Reference to this dataset:**

Brown, G.G., James, S.W., Csuzdi, C., Lapied, E., Decaëns, T., Reynolds, J.W., Mısırlıoğlu, M., Stovanić, M., Trakić, T., Sekulić, J., Phillips, H., Cameron, E. (2023)

A checklist of megadrile earthworm (Annelida: Clitellata) species and subspecies of the world.

Zenodo, <https://doi.org/10.5281/zenodo.7301848>

**Reference to original manuscript:**

Mısırlıoğlu, M., Reynolds, J.W., Stovanić, M, Trakić, T, Sekulić, J., James, S.W., Csuzdi, C., Decaëns, T., Lapied, E., Phillips, H.R.P., Cameron, E.K, Brown, G.G. (2023)

Earthworms (Clitellata, Megadrili) of the world: an updated checklist of valid species and families, with notes on their distribution.

Zootaxa. <https://doi.org/10.11646/zootaxa.5255.1.33>

# Data import

iNaturalist Taxonomy DarwinCore Archive:

```{r}
#| label: import-INAT
#| message: FALSE

INAT_raw <- read_csv(
  here("data", "raw", "inaturalist-taxonomy.dwca", "taxa.csv"))
```

WoRMS Crassiclitellata taxon list:

```{r}
#| label: import-WORMS

WORMS_raw <- read_xlsx(
  here("data", "raw", "Crassiclitellata_18-05-2026.xlsx"))
```

Megadrile earthworm (Annelida: Clitellata) species checklist:

```{r}
#| label: import-WME

WME_raw <- read_xlsx(
  here("data", "raw", "World-megadrile-earthworm-species-checklist-2023-Zenodo-v4.xlsx"), 
  sheet = "Species")
```

# Data cleaning

Simplify the raw data by retaining the following columns:

- family

- genus

- specificEpithet

- species\_*DATABASE*

- author (when available)

## INAT

```{r}
#| label: clean-INAT

INAT <- INAT_raw %>% 
  filter(taxonRank == "species",
         order == "Crassiclitellata") %>% 
  select(family,
         genus,
         specificEpithet) %>% 
  mutate(species_INAT = paste(genus, specificEpithet)) %>% 
  arrange(family,
          genus,
          specificEpithet)
```

Duplicate check:

```{r}
#| label: duplicates-INAT

INAT %>% count(species_INAT) %>% filter(n > 1)
```

No duplicates found in the iNaturalist taxonomy dataset.

## WORMS

The raw dataset includes the following status:

```{r}
#| label: status-WORMS

unique(WORMS_raw$Status)
```

(at)thirty_legs commented in the iNaturalist flag:

"It also seems like iNaturalist did not import **some** species that have the uncertain/unassessed status on WoRMS, leading to further holes in the iNat taxonomy."

Since the comment implies that only "some" species with these status were not imported,

I will only remove the "unaccepted" species from the database.

```{r}
#| label: clean-WORMS

WORMS <- WORMS_raw %>%
  filter(RankName == "Species", 
         Status != "unaccepted") %>% 
  select(family = Family,
         genus = Genus,
         specificEpithet = Species,
         author = Authority,
         species_WORMS = ScientificName) %>% 
  distinct(species_WORMS,
           author,
           .keep_all = TRUE) %>%
  arrange(family,
          genus,
          specificEpithet)
```

Duplicate check:

```{r}
#| label: duplicates-WORMS

WORMS %>%
  count(species_WORMS) %>%
  filter(n > 1)
```

18 duplicate species detected in the WoRMS taxon list

## WME

```{r}
#| label: clean-WME

WME <- WME_raw %>%
  select(family = Family,
         genus = Genus,
         specificEpithet = species,
         author = taxon_author) %>%
  mutate(species_WME = paste(genus, specificEpithet)) %>% 
  distinct(species_WME, #mantener species=subspecies!!!!!!!!!!
           .keep_all = TRUE) %>% 
   arrange(family,
          genus,
          specificEpithet)
```

Duplicate check:

```{r}
#| label: duplicates-WME

WME %>%
  count(species_WME) %>%
  filter(n > 1)
```

No duplicates found in the WME taxonomy dataset.

# Results

## Species total

### INAT

Species count:

```{r}
#| label: species-INAT

sppINAT <- n_distinct(INAT$species_INAT)
```

Species total = `r sppINAT`.

Only 201 spp currently have Research Grade observations on iNaturalist (24-05-2026).

Family summary:

```{r}
#| label: sum-INAT

INAT |> count(family)
```

### WORMS

Count species:

```{r}
#| label: species-WORMS

sppWORMS <- n_distinct(WORMS$species_WORMS)
```

Species total = `r sppWORMS` - duplicados = 5537. !!!!!!!!!!!!!!!!!!!!!!

Family summary:

```{r}
#| label: sum-WORMS

WORMS |> count(family)
```

### WME

Count species:

```{r}
#| label: species-WME

sppWME <- n_distinct(WME$species_WME)
```

Species total = `r sppWME`.

Family summary

```{r}
#| label: sum-WME

WME |> count(family)
```

## Alignment

### INAT vs WORMS

How many iNaturalist species do not match the World Register of Marine Species?

```{r}
#| label: INAT-vs-WORMS

unmatchedWORMS <- anti_join(INAT, 
                            WORMS, 
                            by = c("species_INAT" = "species_WORMS"))
nrow(unmatchedWORMS)
```

219 unmatched spp (34.9%). !!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!

```{r}
coverage_WORMS <- 100 * (
  (sppINAT - nrow(unmatchedWORMS)) / sppINAT
)
```

```{r}
#| label: export-unmatchedWORMS

write_csv(
  unmatchedWORMS,
  here("outputs", "unmatched_WORMS.csv"))
```

### INAT vs WME

How many iNaturalist species do not match the World Megadrile Earthworm database?

```{r}
#| label: INAT-vs-WME

unmatchedWME <- anti_join(INAT, 
                          WME, 
                          by = c("species_INAT" = "species_WME")) 
nrow(unmatchedWME)
```

72 unmatched spp (11.5%). !!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!

```{r}
coverage_WME <- 100 * (
  (sppINAT - nrow(unmatchedWME)) / sppINAT
)
```

```{r}
#| label: export-unmatchedWME

write_csv(
  unmatchedWME,
  here("outputs", "unmatched_WME.csv"))
```

### Combined coverage

Are all active iNaturalist earthworm species represented in at least one external framework?

```{r}
#| label: WORMS+WME

WORMS_WME <- bind_rows(
  WORMS %>% rename(species = species_WORMS),
  WME %>% rename(species = species_WME)
)
```

```{r}
#| label: INATvsWORMS+WME

unmatchedALL <- anti_join(INAT, 
                          WORMS_WME, 
                          by = c("species_INAT" = "species"))
nrow(unmatchedALL)
```

```{r}
#| label: export-unmatched

write_csv(
  unmatchedALL,
  here("outputs", "unmatched_ALL.csv"))
```
