---
title: "species-comparison"
author: "Cenaida Largo"
format: html
editor: visual
---

# Overview

This script explores the alignment of the iNaturalist earthworm (Crassiclitellata) taxonomy with:

- **WoRMS** (World Register of Marine Species), the current taxonomy source used by iNaturalist.

- **Misırlıoğlu et al.(2023)**, an updated global checklist of valid earthworm species.

Subspecies are excluded from this analysis.

# Setup

Package versions are tracked with `renv`.

```{r}
#| label: setup-packages
#| message: false
#| warning: false

library(here)     #Project relative file paths
library(readr)    #Import data files into R
library(readxl)   #Import excel files into R
library(dplyr)    #Manipulating data
library(tidyr)    #Tidy data
library(stringr)  #Manipulating text
library(ggplot2)  #Data visualisations
```

# Data import

iNaturalist Taxonomy DarwinCore Archive

```{r}
#| label: import-INAT
#| message: FALSE

INAT_raw <- read_csv(
  here("data", "raw", "inaturalist-taxonomy.dwca", "taxa.csv"))
```

WoRMS Crassiclitellata taxon list

```{r}
#| label: import-WORMS

WORMS_raw <- read_xlsx(
  here("data", "raw", "Crassiclitellata_18-05-2026.xlsx"))
```

Megadrile earthworm (Annelida: Clitellata) species checklist

```{r}
#| label: import-WME

WME_raw <- read_xlsx(
  here("data", "raw", "World-megadrile-earthworm-species-checklist-2023-Zenodo-v4.xlsx"), 
  sheet = "Species")
```

# Data cleaning

Simplify the raw data by retaining the following columns:

- family

- genus

- specificEpithet

- species\_*DATABASE*

- author (when available)

## INAT

```{r}
#| label: clean-INAT

INAT <- INAT_raw %>% 
  filter(taxonRank == "species",
         order == "Crassiclitellata") %>% 
  select(family,
         genus,
         specificEpithet) %>% 
  mutate(species_INAT = paste(genus, specificEpithet)) %>% 
  arrange(family,
          genus,
          specificEpithet)
```

Duplicate check

```{r}
#| label: duplicates-INAT
#| message: FALSE

INAT_dup <- nrow(INAT %>%
                   count(species_INAT) %>%
                   filter(n > 1))
```

`r INAT_dup` duplicate species detected in the iNaturalist taxonomy dataset.

## WORMS

The raw dataset includes the following status categories:

```{r}
#| label: status-WORMS

unique(WORMS_raw$Status)
```

(at)thirty_legs commented in the iNaturalist flag:

"It also seems like iNaturalist did not import **some** species that have the uncertain/unassessed status on WoRMS, leading to further holes in the iNat taxonomy."

Since the comment implies that only "some" species with these status were not imported,

I will only remove the "unaccepted" species from the database.

```{r}
#| label: clean-WORMS

WORMS <- WORMS_raw %>%
  filter(RankName == "Species", 
         Status != "unaccepted") %>% 
  select(family = Family,
         genus = Genus,
         specificEpithet = Species,
         author = Authority,
         species_WORMS = ScientificName) %>% 
  distinct(species_WORMS,
           author,
           .keep_all = TRUE) %>%
  arrange(family,
          genus,
          specificEpithet)
```

Duplicate check

```{r}
#| label: duplicates-WORMS

WORMS_dup <- nrow(WORMS %>%
  count(species_WORMS) %>%
  filter(n > 1))
```

`r WORMS_dup` duplicate species detected in the WoRMS taxon list.

## WME

```{r}
#| label: clean-WME

WME <- WME_raw %>%
  select(family = Family,
         genus = Genus,
         specificEpithet = species,
         author = taxon_author) %>%
  mutate(species_WME = paste(genus, specificEpithet)) %>% 
  distinct(species_WME, #mantener species=subspecies!!!!!!!!!!
           .keep_all = TRUE) %>% 
   arrange(family,
          genus,
          specificEpithet)
```

Duplicate check:

```{r}
#| label: duplicates-WME

WME_dup <- nrow(WME %>%
  count(species_WME) %>%
  filter(n > 1))
```

`r WME_dup` duplicate species detected in the WME taxonomy dataset.

# Results

## Species total

### INAT

Species count:

```{r}
#| label: species-INAT

sppINAT <- n_distinct(INAT$species_INAT)
```

Species total = `r sppINAT`.

Only 201/628 spp currently (24-05-2026) have Research Grade (RG) observations on iNaturalist.

### WORMS

Species count

```{r}
#| label: species-WORMS

sppWORMS <- n_distinct(WORMS$species_WORMS)
```

Species total = `r sppWORMS`

### WME

Species count

```{r}
#| label: species-WME

sppWME <- n_distinct(WME$species_WME)
```

Species total = `r sppWME`.

## Alignment

### INAT vs WORMS

How many iNaturalist species do not match the World Register of Marine Species?

```{r}
#| label: INAT-vs-WORMS

unmatchedWORMS <- anti_join(INAT, 
                            WORMS, 
                            by = c("species_INAT" = "species_WORMS"))

n_unmatchedWORMS <- nrow(unmatchedWORMS)

write_csv(unmatchedWORMS, here("outputs", "unmatched_WORMS.csv"))
```

`r n_unmatchedWORMS` unmatched spp (`r sprintf("%.2f%%", (n_unmatchedWORMS / sppINAT) * 100)`).

### INAT vs WME

How many iNaturalist species do not match the World Megadrile Earthworm database?

```{r}
#| label: INAT-vs-WME

unmatchedWME <- anti_join(INAT, 
                          WME, 
                          by = c("species_INAT" = "species_WME")) 

n_unmatchedWME <- nrow(unmatchedWME)

write_csv(unmatchedWME, here("outputs", "unmatched_WME.csv"))
```

`r n_unmatchedWME` unmatched spp (`r sprintf("%.2f%%", (n_unmatchedWME / sppINAT) * 100)`).

### Combined coverage

Are all active iNaturalist earthworm species represented in at least one external framework?

```{r}
#| label: WORMS+WME

WORMS_WME <- bind_rows(
  WORMS %>% rename(species = species_WORMS),
  WME %>% rename(species = species_WME))

unmatchedALL <- anti_join(INAT, 
                          WORMS_WME, 
                          by = c("species_INAT" = "species"))

n_unmatchedALL <- nrow(unmatchedALL)

write_csv(unmatchedALL, here("outputs", "unmatched_ALL.csv"))
```

`r n_unmatchedALL` unmatched spp (`r sprintf("%.2f%%", (n_unmatchedALL / sppINAT) * 100)`).

## Data visualization

Coverage of iNaturalist earthworm species

```{r}
#| label: stacked bar chart
coverage <- tibble(
  source = c("WoRMS", "WME"),
  matched = c(
    sppINAT - n_unmatchedWORMS,
    sppINAT - n_unmatchedWME
  ),
  unmatched = c(
    n_unmatchedWORMS,
    n_unmatchedWME
  )
)

coverage_long <- coverage %>%
  pivot_longer(
    c(matched, unmatched),
    names_to = "status",
    values_to = "species"
  ) %>%
  mutate(
    status = factor(status,
                    levels = c("unmatched", "matched"))
  )

fig_coverage <- ggplot(coverage_long,
       aes(source, species, fill = status)) +
  geom_col(
    width = 0.7
    ) +
  coord_flip() +
  scale_fill_manual(
    values = c(
      unmatched = "#4D4D4D",
      matched = "#74AC00"
    ),
    labels = c(
      unmatched = "Not matched",
      matched = "Matched"
    )
  ) +
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.05))
  ) +
  labs(
    x = NULL,
    y = "iNaturalist earthworm species",
    fill = NULL
  ) +
  theme_classic(base_size = 12) +
  theme(
    legend.position = "top",
    legend.justification = "left",
    axis.title.y = element_blank(),
    axis.title.x = element_text(
      margin = margin(t = 10)
    ),
    axis.text = element_text(color = "black"),
    legend.text = element_text(size = 11)
  ) +
  geom_text(
  aes(label = ifelse(status == "unmatched",
                     paste0(round(species / sppINAT * 100, 1), "%"),
                     "")),
  position = position_stack(vjust = 0.5),
  color = "white",
  size = 3.5
)
```

```{r}
#| label: figure upload
ggsave(
  filename = here("outputs", "coverage_INAT.png"),
  plot = fig_coverage,
  width = 7,
  height = 5,
  units = "in",
  dpi = 600
)
```
