---
title: "species-comparison"
author: "Cenaida Largo"
format: html
editor: visual
---

# Overview

This script explores the alignment of the iNaturalist earthworm (Crassiclitellata) taxonomy with:

- **WoRMS** (World Register of Marine Species), the current taxonomy source used by iNaturalist.

- **Misırlıoğlu et al. (2023)**, an updated global checklist of valid earthworm species.

Subspecies are excluded from this analysis.

## Setup

Package versions are tracked with \`renv\`.

```{r}
#| label: setup-packages
#| message: false
#| warning: false

library(here)     #Project relative file paths
library(readr)    #Import data files into R
library(readxl)   #Import excel files into R
library(dplyr)    #Manipulating data
library(stringr)  #Manipulating text
```

## Data sources

### iNaturalist

Download all “active” names from the **iNaturalist Taxonomy DarwinCore Archive** at

<https://www.inaturalist.org/pages/developers>

Updated monthly.

### WoRMS

**WoRMS taxon list** downloads are available at

<https://www.marinespecies.org/aphia.php?p=download&what=taxlist>

Because the web interface limits exports to 1000 rows, a custom export was requested using:

<https://www.marinespecies.org/usersrequest.php>

Requested fields:

- ScientificName

- Authority

- taxonRank

- taxonomicStatus

- Classification

- Accepted taxon information.

Make sure to change the environment field from “marine = yes” to “marine = any”.

Contact: info(at)marinespecies.org

### **Misirlioğlu 2023**

**Reference to this dataset:**

Brown, G.G., James, S.W., Csuzdi, C., Lapied, E., Decaëns, T., Reynolds, J.W., Mısırlıoğlu, M., Stovanić, M., Trakić, T., Sekulić, J., Phillips, H., Cameron, E. (2023)

A checklist of megadrile earthworm (Annelida: Clitellata) species and subspecies of the world.

Zenodo, <https://doi.org/10.5281/zenodo.7301848>

**Reference to original manuscript:**

Mısırlıoğlu, M., Reynolds, J.W., Stovanić, M, Trakić, T, Sekulić, J., James, S.W., Csuzdi, C., Decaëns, T., Lapied, E., Phillips, H.R.P., Cameron, E.K, Brown, G.G. (2023)

Earthworms (Clitellata, Megadrili) of the world: an updated checklist of valid species and families, with notes on their distribution.

Zootaxa. <https://doi.org/10.11646/zootaxa.5255.1.33>

# Data import

iNaturalist Taxonomy DarwinCore Archive:

```{r}
#| label: import-INAT

INAT_raw <- read_csv(
  here("data", "raw", "inaturalist-taxonomy.dwca", "taxa.csv"))
```

WoRMS Crassiclitellata taxon list:

```{r}
#| label: import-WORMS

WORMS_raw <- read_xlsx(
  here("data", "raw", "Crassiclitellata_18-05-2026.xlsx"))
```

Megadrile earthworm (Annelida: Clitellata) species checklist:

```{r}
#| label: import-WME

WME_raw <- read_xlsx(
  here("data", "raw", "World-megadrile-earthworm-species-checklist-2023-Zenodo-v4.xlsx"), 
  sheet = "Species")
```

# Data cleaning

Simplify the raw data by retaining the following columns:

- family

- genus

- specificEpithet

- species_database

- author (when available)

## INAT

```{r}
#| label: clean-INAT

INAT <- INAT_raw %>% 
  filter(taxonRank == "species",
         order == "Crassiclitellata") %>% 
  select(family,
         genus,
         specificEpithet) %>% 
  mutate(species_INAT = paste(genus, specificEpithet)) %>% 
  arrange(family,
          genus,
          specificEpithet)
```

## WORMS

```{r}
#| label: clean-WORMS

WORMS <- WORMS_raw %>%
  filter(RankName == "Species",
         Status != "unaccepted") %>%
  select(family = Family,
         genus = Genus,
         specificEpithet = Species,
         author = Authority,
         species_WORMS = ScientificName) %>% 
  distinct(species_WORMS,
           author,
           .keep_all = TRUE) %>%
  arrange(family,
          genus,
          specificEpithet)
```

## WME

```{r}
#| label: clean-WME

WME <- WME_raw %>%
  select(family = Family,
         genus = Genus,
         specificEpithet = species,
         author = taxon_author) %>%
  mutate(species_WME = paste(genus, specificEpithet)) %>% 
  distinct(species_WME,
           .keep_all = TRUE) %>% 
   arrange(family,
          genus,
          specificEpithet)
```
